Dependently Typed Multi-Stage Calculus\thanks{Supported by organization x.}

### 2019-05-15 main.tex
Making references.

\newcommand{\red}[1]{\textcolor{red}{ #1 }}
\newcommand{\blue}[1]{\textcolor{blue}{ #1 }}

@@

\begin{document}
%
\title{Dependent Typed Multi-Stage Calculus\thanks{Supported by organization x.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Akira Kawata\inst{1} \and
Atsushi Igarashi\inst{2}\orcidID{0000-0002-5143-9764}}
%
\authorrunning{A. Kawata, A. Igarashi}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Graduate School of Informatics, Kyoto University, Kyoto, Japan\\
\email{akira@fos.kuis.kyoto-u.ac.jp} \and
\email{igarashi@kuis.kyoto-u.ac.jp}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}

We develop yet another typed multi-stage calculus \LMD.
It extends Hanada and Igarashi's \LTP with dependent type.

@@

Dependent types is types dependent on values. 
Vectors with lengths is famous example of dependent types and they enable us to omit boundary checking.

@@

It has a simple, substitution-based full-reduction semantics and enjoys basic properties of subject reduction, confluence, strong normalization and progress.

The main technical points of this paper is how to deal with Cross Stage Persistence of multi-stage calculuses in a dependent type system.
CSP is a function of a multi-stage calculus which allows to use a value in quoted code.

The first point is CSP in kinding rules.
$\lambda^{\TW\%}$ includes kinding rules because dependent types need them
but the relationship between kinds and stages was't clear.

The second point is CSP in equivalence rules.
Prameters of dependent types may contain cross-staed terms.
We give reasonable equivalence rules to handle them.

@@

\begin{align*}
    \textrm{Terms} && M,N,L,O,P & ::= x \mid \lambda x:\tau.M\ \mid M\ M \mid \TB_\alpha M 
    \mid \TBL_\alpha M \mid \Lambda\alpha.M \mid M\ \epsilon \mid \%_\alpha M\\ 
    \textrm{Types} && \tau,\sigma,\rho,\pi,\xi & ::= X \mid \Pi x:\tau.\tau \mid \tau\ M \mid \TW_{\alpha} M \mid \F\alpha.\tau \\
    \textrm{Kinds} && K,J,I,H,G & ::= * \mid \Pi x:\tau.K\\
    \textrm{Contexts} && \Gamma & ::= \phi \mid 
    \Gamma,x:\tau@A\ (x\not\in\textrm{FV}(\G)) \mid 
    \Gamma,X::K@A\ (X\not\in \textrm{FV}(\G))\\
    \textrm{Transition variables} && & \alpha,\beta,\gamma,... \\
    \textrm{Transition} && & A,B,C,... \\
    \textrm{Variables} && & x,y,z,... \\
    \textrm{Type variables} && & X,Y,Z,... \\
\end{align*}

@@

\begin{theorem}[Substituition Lemma for Variables of Terms, Types or Kinds and Their Equivalence]
    \begin{flalign*}
        \text{If\ } \G,z:\xi@B \V M:\tau@A \text{\ and\ } \G\V P:\xi@B
        &\text{\ then\ } \G\V M[z \mapsto P]:\tau[z \mapsto P]@A.&\\
        \text{If\ } \G,z:\xi@B \V \tau::K@A \text{\ and\ } \G\V P:\xi@B
        &\text{\ then\ } \G\V \tau[z \mapsto P]::K[z \mapsto P]@A.&\\
        \text{If\ } \G,z:\xi@B \V K\iskind@A \text{\ and\ } \G\V P:\xi@B
        &\text{\ then\ } \G\V K[z \mapsto P] \iskind @A.&\\
        \text{If\ } \G,z:\xi@B \V M\E N : \tau@A \text{\ and\ } \G\V P:\xi@B
        &\text{\ then\ } \G\V M[z \mapsto P]\E N[z \mapsto P] : \tau[z \mapsto P]@A.&\\
        \text{If\ } \G,z:\xi@B \V \tau\E \sigma : K@A \text{\ and\ } \G\V P:\xi@B
        &\text{\ then\ } \G\V \tau[z \mapsto P]\E \sigma[z \mapsto P] : K[z \mapsto P]@A.&\\
        \text{If\ } \G,z:\xi@B \V K\E J@A \text{\ and\ } \G\V P:\xi@B
        &\text{\ then\ } \G\V K[z \mapsto P]\E J[z \mapsto P]@A.&
    \end{flalign*}    
\end{theorem}

\begin{theorem}[Stage Substituition Lemma for Variables of Terms, Types or Kinds and Their Equivalence]
    \begin{flalign*}
        \text{If\ } \G \V M:\tau@A
        &\text{\ then\ } \G[\beta \mapsto \epsilon]\V M[\beta \mapsto \epsilon]:\tau[\beta \mapsto \epsilon]@A[\beta \mapsto \epsilon].&\\
        \text{If\ } \G \V \tau::K@A
        &\text{\ then\ } \G[\beta \mapsto \epsilon]\V \tau[\beta \mapsto \epsilon]::K[\beta \mapsto \epsilon]@A[\beta \mapsto \epsilon].&\\
        \text{If\ } \G \V K\iskind@A
        &\text{\ then\ } \G[\beta \mapsto \epsilon]\V K[\beta \mapsto \epsilon] \iskind@A[\beta \mapsto \epsilon].&\\
        \text{If\ } \G \V M\E N : \tau@A
        &\text{\ then\ } \G[\beta \mapsto \epsilon]\V M[\beta \mapsto \epsilon]\E N[\beta \mapsto \epsilon] : \tau[\beta \mapsto \epsilon] @A[\beta \mapsto \epsilon].&\\
        \text{If\ } \G \V \tau\E \sigma : K@A
        &\text{\ then\ } \G[\beta \mapsto \epsilon]\V \tau[\beta \mapsto \epsilon]\E \sigma[\beta \mapsto \epsilon] : K[\beta \mapsto \epsilon]@A[\beta \mapsto \epsilon].&\\
        \text{If\ } \G \V K\E J@A
        &\text{\ then\ } \G[\beta \mapsto \epsilon]\V K[\beta \mapsto \epsilon]\E J[\beta \mapsto \epsilon]@A[\beta \mapsto \epsilon].&
    \end{flalign*}
\end{theorem}

\begin{theorem}[Preservation for term on $\beta$ reduction]
    If $\G\V M:\tau@A$ and $M \longrightarrow_{\beta} M'$, then $\G\V M':\tau@A$\\
\end{theorem}

\begin{theorem}[Preservation for term on $\TBL\TB$ reduction]
    If $\G\V M:\tau@A$ and $M\longrightarrow_\blacklozenge N$, then $\G\V N:\tau@A$\\
\end{theorem}

\begin{theorem}[Preservation for term on $\Lambda$ reduction]
    If $\G\V M:\tau@A$ and $M \longrightarrow_{\Lambda} N$, then $\G\V N:\tau@A$.
\end{theorem}

@@

\begin{definition}[Values]
$A \neq \epsilon$\\
\begin{align*}
    \textrm{Values} && v^\epsilon \in V^\epsilon & ::= \lambda x:\tau.M \mid\ \TB_\alpha v^\alpha \mid \Lambda\alpha.v^\epsilon & \\
                    && v^A \in V^A & ::= x \mid \lambda x:\tau.v^A \mid v^A\ v^A \mid\ \TB_\alpha v^{A\alpha} 
                                           \mid \Lambda\alpha.v^A \mid v^A\ \epsilon &\\
                                    &&& \quad\   \mid\ \TBL_\alpha v^{A'} (\text{if } A'\alpha = A \text{ and } A' \neq \epsilon) & \\
                                    &&& \quad\   \mid \%_\alpha v^{A'} (\text{if } A'\alpha = A) & \\
    \textrm{Redexes} && R^\epsilon & ::= (\lambda x:\tau.M)\ v^\epsilon \mid (\Lambda\alpha.v^\epsilon)\ \epsilon & \\
                     && R^\alpha & ::=\ \TBL_\alpha \TB_\alpha M & \\
\end{align*}
\end{definition}

\begin{theorem}[Progress]
    If $x:\tau@\epsilon \notin \G$ and $\G \V M : \tau @ A$ then $ M \in V^A $ or $\exists M'$ such that $M \longrightarrow M'$.
\end{theorem}

@@

>>>>

\newcommand{\red}[1]{\textcolor{red}{#1 }}
\newcommand{\blue}[1]{\textcolor{blue}{#1 }}

@@

\begin{document}
%
\title{Dependently Typed Multi-Stage Calculus\thanks{Supported by organization x.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Akira Kawata\inst{1} \and
Atsushi Igarashi\inst{2}\orcidID{0000-0002-5143-9764}}
%
\authorrunning{A. Kawata, A. Igarashi}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Graduate School of Informatics, Kyoto University, Kyoto, Japan\\
\email{akira@fos.kuis.kyoto-u.ac.jp} \and
\email{igarashi@kuis.kyoto-u.ac.jp}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}

We develop yet another typed multi-stage calculus \LMD.
It extends Hanada and Igarashi's \LTP with dependent types.

@@

Dependent types are types dependent on values. 
Vectors with lengths are a famous example of dependent types and they enable us to omit boundary checking.

@@

It has a simple, substitution-based full-reduction semantics and enjoys basic properties of subject reduction, confluence, strong normalization, and progress.

The main technical points of this paper are how to deal with Cross Stage Persistence of multi-stage calculuses in a dependent type system.
CSP is a function of a multi-stage calculus which allows using a value in quoted code.

The first point is CSP in kinding rules.
$\lambda^{\TW\%}$ includes kinding rules because dependent types need them
but the relationship between kinds and stages wasn't clear.

The second point is CSP in equivalence rules.
Parameters of dependent types may contain cross-staged terms.
We give reasonable equivalence rules to handle them.

@@

\begin{align*}
    \textrm{Terms} && M,N,L,O,P & ::= x \mid \lambda x:\tau.M\ \mid M\ M \mid \TB_\alpha M 
    \mid \TBL_\alpha M \mid \Lambda\alpha.M \mid M\ \epsilon \mid \%_\alpha M\\ 
    \textrm{Types} && \tau,\sigma,\rho,\pi,\xi & ::= X \mid \Pi x:\tau.\tau \mid \tau\ M \mid \TW_{\alpha} M \mid \F\alpha.\tau \\
    \textrm{Kinds} && K,J,I,H,G & ::= * \mid \Pi x:\tau.K\\
    \textrm{Contexts} && \Gamma & ::= \phi \mid 
    \Gamma,x:\tau @A\ (x\not\in\textrm{FV}(\G)) \mid 
    \Gamma,X::K @A\ (X\not\in \textrm{FV}(\G))\\
    \textrm{Transition variables} && & \alpha,\beta,\gamma, \dots \\
    \textrm{Transition} && & A,B,C,\dots \\
    \textrm{Variables} && & x,y,z,\dots \\
    \textrm{Type variables} && & X,Y,Z,\dots \\
\end{align*}

@@

\begin{theorem}[Substituition Lemma for Variables of Terms, Types or Kinds and Their Equivalence]
    \begin{flalign*}
        \text{If\ } \G,z:\xi @B \V M:\tau @A \text{\ and\ } \G\V P:\xi @B
        &\text{\ then\ } \G\V M[z \mapsto P]:\tau[z \mapsto P] @A.&\\
        \text{If\ } \G,z:\xi @B \V \tau::K @A \text{\ and\ } \G\V P:\xi @B
        &\text{\ then\ } \G\V \tau[z \mapsto P]::K[z \mapsto P] @A.&\\
        \text{If\ } \G,z:\xi @B \V K\iskind @A \text{\ and\ } \G\V P:\xi @B
        &\text{\ then\ } \G\V K[z \mapsto P] \iskind  @A.&\\
        \text{If\ } \G,z:\xi @B \V M\E N : \tau @A \text{\ and\ } \G\V P:\xi @B
        &\text{\ then\ } \G\V M[z \mapsto P]\E N[z \mapsto P] : \tau[z \mapsto P] @A.&\\
        \text{If\ } \G,z:\xi @B \V \tau\E \sigma : K @A \text{\ and\ } \G\V P:\xi @B
        &\text{\ then\ } \G\V \tau[z \mapsto P]\E \sigma[z \mapsto P] : K[z \mapsto P] @A.&\\
        \text{If\ } \G,z:\xi @B \V K\E J @A \text{\ and\ } \G\V P:\xi @B
        &\text{\ then\ } \G\V K[z \mapsto P]\E J[z \mapsto P] @A.&
    \end{flalign*}    
\end{theorem}

\begin{theorem}[Stage Substituition Lemma for Variables of Terms, Types or Kinds and Their Equivalence]
    \begin{flalign*}
        \text{If\ } \G \V M:\tau @A
        &\text{\ then\ } \G[\beta \mapsto \epsilon]\V M[\beta \mapsto \epsilon]:\tau[\beta \mapsto \epsilon] @A[\beta \mapsto \epsilon].&\\
        \text{If\ } \G \V \tau::K @A
        &\text{\ then\ } \G[\beta \mapsto \epsilon]\V \tau[\beta \mapsto \epsilon]::K[\beta \mapsto \epsilon] @A[\beta \mapsto \epsilon].&\\
        \text{If\ } \G \V K\iskind @A
        &\text{\ then\ } \G[\beta \mapsto \epsilon]\V K[\beta \mapsto \epsilon] \iskind @A[\beta \mapsto \epsilon].&\\
        \text{If\ } \G \V M\E N : \tau @A
        &\text{\ then\ } \G[\beta \mapsto \epsilon]\V M[\beta \mapsto \epsilon]\E N[\beta \mapsto \epsilon] : \tau[\beta \mapsto \epsilon]  @A[\beta \mapsto \epsilon].&\\
        \text{If\ } \G \V \tau\E \sigma : K @A
        &\text{\ then\ } \G[\beta \mapsto \epsilon]\V \tau[\beta \mapsto \epsilon]\E \sigma[\beta \mapsto \epsilon] : K[\beta \mapsto \epsilon] @A[\beta \mapsto \epsilon].&\\
        \text{If\ } \G \V K\E J @A
        &\text{\ then\ } \G[\beta \mapsto \epsilon]\V K[\beta \mapsto \epsilon]\E J[\beta \mapsto \epsilon] @A[\beta \mapsto \epsilon].&
    \end{flalign*}
\end{theorem}

\begin{theorem}[Preservation for term on $\beta$ reduction]
    If $\G\V M:\tau @A$ and $M \longrightarrow_{\beta} M'$, then $\G\V M':\tau @A$\\
\end{theorem}

\begin{theorem}[Preservation for term on $\TBL\TB$ reduction]
    If $\G\V M:\tau @A$ and $M\longrightarrow_\blacklozenge N$, then $\G\V N:\tau @A$\\
\end{theorem}

\begin{theorem}[Preservation for term on $\Lambda$ reduction]
    If $\G\V M:\tau @A$ and $M \longrightarrow_{\Lambda} N$, then $\G\V N:\tau @A$.
\end{theorem}

@@

\begin{definition}[Values]
$A \neq \epsilon$\\
\begin{align*}
    \textrm{Values} && v^\epsilon \in V^\epsilon & ::= \lambda x:\tau.M \mid\ \TB_\alpha v^\alpha \mid \Lambda\alpha.v^\epsilon & \\
                    && v^A \in V^A & ::= x \mid \lambda x:\tau.v^A \mid v^A\ v^A \mid\ \TB_\alpha v^{A\alpha} 
                                           \mid \Lambda\alpha.v^A \mid v^A\ \epsilon &\\
                                    &&& \quad\   \mid\ \TBL_\alpha v^{A'} (\text{if } A'\alpha = A \text{and } A' \neq \epsilon) & \\
                                    &&& \quad\   \mid \%_\alpha v^{A'} (\text{if } A'\alpha = A) & \\
    \textrm{Redexes} && R^\epsilon & ::= (\lambda x:\tau.M)\ v^\epsilon \mid (\Lambda\alpha.v^\epsilon)\ \epsilon & \\
                     && R^\alpha & ::=\ \TBL_\alpha \TB_\alpha M & \\
\end{align*}
\end{definition}

\begin{theorem}[Progress]
    If $x:\tau @\epsilon \notin \G$ and $\G \V M : \tau  @ A$ then $ M \in V^A $ or $\exists M'$ such that $M \longrightarrow M'$.
\end{theorem}

@@

\begin{itemize}
   \item On Cross-Stage Persistence in Multi-Stage\cite{Hanada2014}
   \item Eliminating Array Bound Checking Through Dependent Types\cite{Xi98}
   \item MetaML and Multi-stage Programming with Explicit Annotations\cite{MetaML}
   \item Idris, a general-purpose dependently typed programming language: Design and implementation\cite{brady2013idris}
   \item A Logical Foundation for Environment Classifiers\cite{Tsukada}
   \item Environment classifiers\cite{taha2003environment}
   \item A framework for defining logics\cite{harper1993framework}
   \item The Design and Implementation of {BER} MetaOCaml - System Description\cite{oleg2014}
   \item Refined Environment Classifiers\cite{kiselyov2016refined}
   \item Staging with control: type-safe multi-stage programming with control operators\cite{oishi2017staging}
\end{itemize}
